The need for accuracy and smoothness in numerical simulations

### 2024-05-08 report/src/draft.tex
Fixed misprints preventing compilation

The GROMACS library is highly optimized in order to reduced the time-to-solution. In the case of the force-field calculation it is common to ignore the interaction between particles that are judged to be sufficiently far away. In practice, this is done by setting individual force-fields to zero outside of a sufficiently large ball. There is more than way to achieve this goal and the online documentation suggest that GROMACS operates with force-fields that are of class $C^1$. Why is this a concern in the context of practical error estimation? Proving that an asymptotic error expansion exist is often an exercise in manipulating Taylor expansions and this is only possible when the functions are sufficiently differentiable. We have designed a sequence of numerical experiment that illustrate this exact point. We consider a system of $m$ identical ions moving in a liquid. They repel each other according to Coulumb's law and they are pulled towards the origo according to Hooke's law. The friction between each ion and the fluid is proportional to its current velocity. This is a simple setup that ensure that particles will ultimate come to rest in a position where all forces balances each other. Can we estimate the value of the electrostatic energy at at specific time $T$? This is the question that we explore in a sequence of 4 scripts. In {\tt iontrap_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$. In {\tt opmtrap_mwe2} we modify the force-field $f = f(r)$ generated by an ion, by retain the exact value of the force-field inside a ball centered on the ion with radius $\rho/2$ and settting it to zero outside the ball. This ensures introduces jump-discontinuites into the simulation. In {\tt iontrap_mwe4} the transition to zero is done smoothly in the region from $0.5 \rho < r < 0.95 r$. It is clear that changing the force-fields is equivalent to physical changes, but can we estimate the discretization error? Figure \ref{}, \ref{}, \ref{} show the evolution of Richardson's fraction for the total kinetic energy at the end of

>>>>

The GROMACS library is highly optimized in order to reduced the time-to-solution. In the case of the force-field calculation it is common to ignore the interaction between particles that are judged to be sufficiently far away. In practice, this is done by setting individual force-fields to zero outside of a sufficiently large ball. There is more than way to achieve this goal and the online documentation suggest that GROMACS operates with force-fields that are of class $C^1$. Why is this a concern in the context of practical error estimation? Proving that an asymptotic error expansion exist is often an exercise in manipulating Taylor expansions and this is only possible when the functions are sufficiently differentiable. We have designed a sequence of numerical experiment that illustrate this exact point. We consider a system of $m$ identical ions moving in a liquid. They repel each other according to Coulumb's law and they are pulled towards the origo according to Hooke's law. The friction between each ion and the fluid is proportional to its current velocity. This is a simple setup that ensure that particles will ultimate come to rest in a position where all forces balances each other. Can we estimate the value of the electrostatic energy at at specific time $T$? This is the question that we explore in a sequence of 4 scripts. In {\tt iontrap\_mwe1} we do not modify the electrostatic force-fields and the $m=4$ ions ultimately form a regular tetrahedron with edge length $\rho$. In {\tt ointrap\_mwe2} we modify the force-field $f = f(r)$ generated by an ion, by retain the exact value of the force-field inside a ball centered on the ion with radius $\rho/2$ and settting it to zero outside the ball. This ensures introduces jump-discontinuites into the simulation. In {\tt iontrap\_mwe4} the transition to zero is done smoothly in the region from $0.5 \rho < r < 0.95 r$. It is clear that changing the force-fields is equivalent to physical changes, but can we estimate the discretization error? Figure \ref{}, \ref{}, \ref{} show the evolution of Richardson's fraction for the total kinetic energy at the end of
